Comparing and Combining Biomarkers as Principle Surrogates for Time-to-Event Clinical Endpoints

### 2014-06-16 Compare_053014_statsmed_revisions.tex
revisions

\runninghead{EE Gabriel and MC Sachs and PB Gilbert}
\title{Comparing and Combining Biomarkers as Principle Surrogates for Time-to-Event Clinical Endpoints}
\author{Erin E. Gabriel\affil{a}\corrauth\
Michael C. Sachs\affil{b}, Peter B. Gilbert\affil{a,c}}
\address{\affilnum{a}Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center\\
\affilnum{b}Kidney Research Institute, University of Washington\\
\affilnum{c}Department of Biostatistics, University of Washington}
\corraddr{egabriel@fhcrc.org}
\begin{abstract}
Principal surrogate endpoints are useful as targets for Phase I and II trials. In many recent trials, multiple post-randomization biomarkers are measured. However, few statistical methods exist for comparison of or combination of biomarkers as principal surrogates and none of these methods to our knowledge utilize time-to-event clinical endpoint information. We propose a Weibull model extension of the semi-parametric estimated maximum likelihood method of \citet{Huang11} that allows for the inclusion of multiple biomarkers in the same risk model as multivariate principal surrogates. We propose several methods for comparing candidate principal surrogates and evaluating multivariate principal surrogates. These include the time-dependent and surrogate-dependent true and false positive fraction, the time-dependent and the integrated standardized total gain and the cumulative distribution function of the risk difference. We illustrate the operating characteristics of our proposed methods in simulations and outline how these statistics can be used to evaluate and compare candidate principal surrogates. We use these methods to investigate candidate surrogates in the Diabetes Control and Complications Trial.
\end{abstract}
\keywords{Causal inference; Comparison of candidate principal stratification; Survival analysis; Accuracy measures; Multivariate principal stratification}
\maketitle

@@

We combine the estimation methods of \citet{Gabriel13} and  \citet{Huang11}  to allow for the evaluation of multivariate principal surrogates of continuous time-to-event clinical endpoints. Much of the literature focuses on the evaluation of univariate biomarkers as principal surrogates for a binary clinical endpoint. \citet{Gabriel13} extended methods for evaluating univariate biomarker as a PS of continuous time-to-event clinical endpoints. \citet{Huang11} introduced a semi-parametric EML method to accommodate evaluation of multivariate candidate surrogates for binary clinical endpoints.  Our proposed method uses a semi-parametric estimated maximum likelihood fit for a Weibull time-to-event model.

@@

The dimension of the marginal-CEP versus candidate PS curve increases with the dimension of the candidate surrogate, making the marginal-CEP difficult to display and interpret for multivariate candidate PS. If instead we consider the clinical outcome within levels of the marginal-CEP, we can define useful curves that are two-dimensional regardless of the number of biomarkers included in the risk model. Our proposed curves include the marginal-CEP-based and time-dependent true positive fraction ($TPF(t|c)$) and false positive fraction ($FPF(t|c)$) as well as the receiver operating characteristic curve ($ROC^{t}(q)$) they define. We also consider the cumulative distribution function (CDF) of the marginal CEP function.

@@

\section{Notation and Criterion for Principal Surrogacy} \label{DEF}
Let $Z$ be the treatment indicator, 0 for control/non-active treatment and 1 for treatment. Let $W$ be a vector of baseline measurements taken prior to randomization. In the principal stratification framework of \citet{Frangakis02} we use potential outcomes, where all post-randomization measures are considered under assignment to either treatment arm for each individual. Let $T_i(z)$ be the potential time from randomization to clinical event for individual $i$ had s/he received treatment $z\in \{0,1\}$. Let $Y(z)$ be the indicator of $T(z)<C(z)$, where $C(z)$ is the potential censoring time and let $X(z)=min(T(z),C(z))$. Let there be $J$ candidate surrogates of interest which is measured prior to the clinical event but after randomization; with $S_{j}(z)$ being the $j$th candidate surrogate under treatment arm $z\in \{0,1\}$. Let $Y_{i}^{t}(z)$ be the indicator that $T_i(z)\leq t$. Let $D_i(t)=Y_{i}^{t}(0)-Y_{i}^{t}(1)$ be the individual treatment effect for subject $i$ on the clinical endpoint at or before time $t$. Let $\{S_{1,i}\ldots,S_{J,i}, T_i, C_i, Y_i, Y_{i}^{t} \}$ denote the observed values of the potential measures for subject $i$.

@@

\citet{Frangakis02} developed the concept of principal surrogacy and defined a PS by a criterion based on joint risk estimands, which conditions on the candidate surrogate under active treatment and under control, $risk_z(s_1,s_0)$. If $risk_1(s_1,s_0)=risk_0(s_1,s_0)$ for all $s_1 = s_0$, S is a principal surrogate; \citet{Gilbert08} called this average causal necessity (ACN). Subsequent work added a second criterion that a good PS should modify the clinical treatment effects over the $(s_1, s_0)$ subgroups, what \citet{Gilbert08} called average causal sufficiency ACS. This can be demonstrated by investigating contrasts of $risk_1(s_1,s_0)$ and $risk_0(s_1,s_0)$, such as the ratio, over the levels of $(s_1, s_0)$.  ACN is a strict but sensible and necessary condition, which guarantees that there is no effect on the clinical outcome when there is no effect on the PS. However, it is clearly not a sufficient condition, as any biomarker in a trial with no clinical treatment effect is likely to satisfy ACN.  The combination of ACS and ACN are often used for evaluation in the current literature, with some works relaxing strict ACN.  ACS also suggests a possible ranking of candidate surrogates based on the level of treatment efficacy modification over the subgroups defined by the surrogate. Strict ACN alone classify candidates as either PS or not PS, with no other indication of quality.

In \citet{Gilbert08}  and most of the recent literature, marginal risk estimands are used instead of joint risk estimands because they are easier to identify and estimate; marginal risk estimands condition on $S(1)$ and marginalize over the distribution of $S(0)$. The concept of ACS is easily extended to marginal risk, such that contrasts of risk over the arms of the trial, such as our CEP of risk difference, vary widely in $s_1$. This can simply be viewed as a subgroup analysis, where subgroups defined by levels of $S(1)$, should strongly modify the clinical treatment effect. Although in some special cases when the candidate surrogate is constant in the control arm, $S(0)=C$ called Constant Biomarker (CB), ACN cannot be verified based on the marginal risks alone \citep{Gilbert08, Gabriel13}.  Often the concept of ACN is relaxed in practice, such that if the CEP of interest is equal to zero when $s_1=0$, ACN is said to be likely. Although this is similar to ACN, this is not the same as ACN unless CB holds and $S(0)=0$. \citet{Gabriel13} outlines some PS candidate constructions that can make CB more likely to hold in a general clinical trial where biomarkers under placebo will tend not to be constant.

For a univariate candidate PS strict or relaxed ACN can be displayed using the joint CEP or the marginal CEP.  Although the concept of ACN is easily extended to a multivariate PS, $risk_1(s_{1,1}, \dots, s_{1,J} ,s_{0,1}, \dots, s_{0,J})=risk_0(s_{1,1}, \dots, s_{1,J} ,s_{0,1} \dots s_{0,J})$ for all $s_{1,j} = s_{0,j}$ with $j \in \{1, \dots, J\}$, it is a more difficult concept to display graphically. The definition does not seem as sensible in the multi-dimensional case. Clearly, we do not want a large amount of clinical treatment effect for a group of subjects with no treatment effect on any of the surrogates. However, it seems like you would want less clinical treatment effect if there was no treatment effect on any of the biomarkers.  We propose to extend the concept of ACN to investigate the ability of the model for risk difference conditional of a set of biomarkers to correctly categorize subjects by their treatment effect at time $t$, $D(t)$. For this purpose, we propose the use of a set of classification accuracy measures, outlined in Sections \ref{CAM} and \ref{CDF}. Like ACN, the classification accuracy measures allow for the quantification of the agreement between the treatment effect on the clinical outcome and the predictive model when there is no treatment effect on any of the biomarkers in the model.  Unlike ACN, the measures allow for quantification of this agreement for all other levels of the biomarkers as well.  These accuracy measures can also be displayed as curves versus the levels of the predictive model to allow for 2-dimensional  illustration of the predictive accuracy of the risk model. We suggest that good agreement between the predictive model containing the set of biomarkers and the clinical treatment efficacy is necessary but not sufficient for a set of biomarkers to have value as a PS. Much like ACN, it is possible to have good agreement between the model prediction and the treatment efficacy simply because there is no efficacy.

ACS for a multivariate PS is wide variation in the CEP over subgroups defined by the set of biomarkers. This is difficult to display graphically due to the increased dimension of the surrogate. We propose the use of functions based on the CEP rather than the set of biomarker values for display of ACS. We propose the CDF of the CEP function described in Section \ref{CDF} for graphical display of multivariate ACS. We also suggest a summary statistic, the standardized total gain, that quantifies both our suggested accuracy concept and ACS in a single measurement. Details for standardized total gain are given in Section \ref{STG}.  As well, if one assumes a parametric model for risk conditional on a set of biomarkers then a test for the minimum amount variation in the CEP over subgroups defined by the biomarkers can be based on the coefficients of the biomarkers in the model. The specifics of this test for our assumed model are given in Section \ref{CRE}.

@@

In order to link these risk estimands to the observable data we must make some assumptions; Assumptions A1-A4 reduce the number of missing potential outcomes and help identify the risk estimands.
\begin{itemize}
\item A1: Stable Unit Treatment Value Assumption (SUTVA) and Consistency 
\item A2: Ignorable Treatment Assignment
\item A3: Equal drop out and individual clinical risk up to time $\tau$: $\min\{T(1), C(1)\} \leq\tau$ if and only if $\min\{T(0), C(0)\} \leq\tau$
\item A4: Random censoring: $T(z) \perp C(z)$ for $z\in \{0,1\}$.
\end{itemize} 
Assumption A1 implies that subjects' potential outcomes are independent of other subjects' outcomes and treatment, SUTVA, and observation of the potential outcome does not change its value, consistency. For example, $(S_i|z_i=0) \equiv S_i(0)$ and this is independent of $(S_j|z_j=1) \equiv S_j(1)$, for $i \neq j$. SUTVA may not hold is some situations, such as cluster randomized vaccine trials where herd immunity may change the treatment effect on all outcomes in the cluster.  Assumption A2 implies that the distribution of the potential outcomes S(z) and Y(z) are independent of the observed treatment assignment. Assumption A2, is generally true in randomized, blinded trials.  Assumption A3 aides in the identification of subjects with min$(T,C) < \tau$ to be excluded from the evaluation cohort without causing bias. Although, A3 is not fully testable, indications of A3 violation are significantly unbalanced events or drop-out over the trial arms prior to $\tau$. Assumption A4 is standard in much of the survival analysis literature and allows for classic censoring models to be used. Assumptions A1-A4 imply that the conditional distribution of $T(z)$ given $\{S_{1}(1)=s_{1,1},\ldots,S_{J}(1)=s_{1,J}, T(1)\geq\tau, T(0)\geq\tau\}$, equals that for the observed outcome T given the observed treatment and potential candidate surrogate measures $\{Z=z, S_{1}(1)=s_{1,1},\ldots,S_{J}(1)=s_{1,J},T\geq\tau\}$ for $z\in \{0,1\}$.

@@

One can test the null hypothesis of no surrogate value or no ACS, $H0_1$, via the estimated model coefficients by a Wald test of the null hypothesis $\gamma_{0j}=\gamma_{1j}=0$ $\forall j \in \{1,\ldots J\}$. If any of the $S_{J}(1)$ associated terms significantly different from zero then there is evidence of variation in the risk difference over the candidate PS.

@@

This likelihood could be solved by direct optimization if we assumed a parametric model for $S_j(1)|W$. However, as proposed in \citet{Pepe91}, consistent estimation of outcome model parameters can be achieved using an estimated distribution of $S_j(1)|W$; this is called estimated maximum likelihood. Following \citet{Huang11}, we assume the semi-parametric location-scale model of \citet{Heagerty99} for $F_{S_{j}(1)|W}$. Considering first a univariate candidate surrogate, the semi-parametric model is defined as $F_{S_{j}(1)|W} \sim F_j[\{s_{1,j} -\mu_j(w)\} / \sigma_j(w)]= F_j(\varsigma_j)$, where $F_j$ is the CDF of the univariate residuals $\varsigma_j$ from fitting the location and scale parameters $\mu_j(w)$ and $\sigma_j(w)$, respectively.  We define $\mu_j(w)$ and $\sigma_j(w)$ as parametric functions of the baseline variable(s) $W$, $\mu_j(w) = \gamma_j'w$ and $\log\{\sigma_j(w)\} = \eta_j'w$. We can estimate these parameters by solving the equations given in \citet{Heagerty99}:
\begin{eqnarray}
&&\sum_{k=1}^{n_V} {w_{k} (s_{(1,j,k)} - \gamma_j' w_{k} )\over \{\exp(\eta_j'w_k)\}^2} =0\\ \label{stuff}
&&\sum_{k=1}^{n_V} {w_{k}\{(s_{(1,j,k)} - \gamma_j' w_{k})^2 - \{\exp(\eta_j'w_k)\}^2\} \over \{\exp(\eta_j'w_k)\}^2}=0,\label{stuff2}
\end{eqnarray}
among those treated subjects with all $S(1)_{(j,k)}$ and $W$ measured. We denote the number of subjects in this validation group by $n_V$.

@@

In this case, sensitivity and specificity are based on unobservable data and cannot be estimated empirically. However, under assumptions A1-A5 plus the additional assumption of monotonicity of treatment effect, we can estimate them. Monotonicity of treatment effect is formally stated as
\begin{itemize}
\item A6: $T(z) \leq T(1-z)$ for one of the treatment arms $z$, $z\in \{0,1\}$.
\end{itemize}
Assumption A6 is needed to identify the classification accuracy measures, but monotonicity does not have to imply that the treatment is never harmful and in some cases the alternative assumption that the treatment never helps is more likely. Assumption A6 is not fully testable, although it can be rejected. Given Assumptions A1-A5, there are clear testable implications of Assumption A6, such as $P(T\leq t|Z=1)\geq P(T\leq t|Z=0)$ for all $t$ if no harm of treatment is assumed, $P(T \leq t|Z=0) \geq P(T\leq t|Z=1)$ for all $t$ if no benefit of treatment is assumed, and no switch in sign of the model-based estimate of risk difference over the range of the observed values of the $S_j(1)$s.

Assumption A6 may be more plausible in settings where the outcome is a specific illness or infection rather than death from any cause. We assume without loss of generality the no-harm of active treatment form of monotonicity heretofore, $T(0) \leq T(1)$, as this is the form of the assumption that is more likely to apply to our motivating example. Assuming A1-A6, $TPF(t|c)$ can be defined as:
\begin{eqnarray*}
TPF(t|c)&=& E\{\Delta(t|S_{1,1}, \dots, S_{1,J})\mathrm{I}_{[\Delta(t|S_{1,1}, \dots, S_{1,J})>c]}\} \over E\{\Delta(t|S_{1,1}, \dots, S_{1,J})\}.
\end{eqnarray*}
The derivation of this form is given in Appendix A of the supplementary materials; both the definition and derivation are adaptations of time-independent curves given in \citet{Huang12b}. The function $FPR(t|c)$ can be defined similarly based on the expectation of $1-\Delta(t|s_{1,1}, \dots, s_{1,J})$. One can estimate these accuracy measures using the plug-in estimators:
\begin{eqnarray*}
\widehat{TPF}(t|c)&=& {\int \hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})\mathrm{I}_{\{\hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})>c\}}d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J}) \over \int \hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J})} \hspace{2mm}\mbox{and},\\
\widehat{FPF}(t|c)&=& {\int 1-\hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})\mathrm{I}_{\{\hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})>c\}}d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J}) \over \int 1-\hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J})}.\\
\end{eqnarray*}
Here $\hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})$ is a maximum likelihood estimate under A1-A6 and using the estimated coefficients from the assumed model A5.

@@

When A6 is observably violated the suggested estimators $FPF(t|c)$ and $TPF(t|c)$ are biased. However, under minor violations of A6 that might not be observable in the sample, the estimators are still reasonable estimates of useful quantities. Greater discussion of this point is given the Appendix A of the supplementary materials. When there are obvious violations of A6, the estimation of $FPF(t|c)$ and $TPF(t|c)$ should not be pursued. When Assumptions A6 is clearly violated we suggest the use of the $\widetilde{STG}$ and $CDF^{t}_{\Delta}(c)$ for comparison and evaluation of candidate surrogates.

@@

\subsection{Integrated Standardized Total Gain} \label{ISTG}
As an alternative to estimating the $STG$ at a fixed time $t$, we can summarize the $STG(t)$ by averaging over the estimated distribution of event times. We define the integrated $STG$ ($ISTG$) as $\int STG(t) \, dF(t)$, where $F(t)$ is the distribution function of the event times. The estimated marginal distribution of event times $\hat{F}(T)$ is calculated by integrating over $Z$ and $S$ in the conditional CDF defined by the assumption A5. Then the plug-in estimator of the integrated $STG$ is the expectation with respect to that distribution:
\[
 \widetilde{STG} = \int \widehat{STG}(t) \, d \hat{F}(t). 
\]
The $\widetilde{STG}$ can be used to summarize the surrogate quality without specifying a fixed time point. By borrowing information across times, it provides a useful, single value estimate for describing and comparing candidate surrogates. The $\widetilde{STG}$ difference between the candidate PS in the same trial can be used in the same way as the difference in $STG(t)$ for inference on comparison.

>>>>

\runninghead{EE Gabriel and MC Sachs and PB Gilbert}
\title{Comparing and Combining Biomarkers as Principle Surrogates for Time-to-Event Clinical Endpoints}
\author{Erin E. Gabriel\affil{a}\corrauth\
Michael C. Sachs\affil{b}, Peter B. Gilbert\affil{a,c}}
\address{\affilnum{a}Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center\\
\affilnum{b}Kidney Research Institute, University of Washington\\
\affilnum{c}Department of Biostatistics, University of Washington}
\corraddr{egabriel@fhcrc.org}
\begin{abstract}
Principal surrogate endpoints are useful as targets for Phase I and II trials. In many recent trials, multiple post-randomization biomarkers are measured. However, few statistical methods exist for comparison of or combination of biomarkers as principal surrogates and none of these methods to our knowledge utilize time-to-event clinical endpoint information. We propose a Weibull model extension of the semi-parametric estimated maximum likelihood method of \citet{Huang11} that allows for the inclusion of multiple biomarkers in the same risk model as multivariate candidate principal surrogates. We propose several methods for comparing candidate principal surrogates and evaluating multivariate principal surrogates. These include the time-dependent and surrogate-dependent true and false positive fraction, the time-dependent and the integrated standardized total gain and the cumulative distribution function of the risk difference. We illustrate the operating characteristics of our proposed methods in simulations and outline how these statistics can be used to evaluate and compare candidate principal surrogates. We use these methods to investigate candidate surrogates in the Diabetes Control and Complications Trial.
\end{abstract}
\keywords{Causal inference; Surrogate endpoint evaluation; Survival analysis; Accuracy measures; Multivariate principal stratification}
\maketitle

@@

We combine the estimation methods of \citet{Gabriel13} and  \citet{Huang11}  to allow for the evaluation of multivariate candidate principal surrogates of continuous time-to-event clinical endpoints. Much of the literature focuses on the evaluation of univariate biomarkers as principal surrogates for a binary clinical endpoint. \citet{Gabriel13} extended methods for evaluating univariate a biomarker as a PS of a continuous time-to-event clinical endpoints. \citet{Huang11} introduced a semi-parametric EML method to accommodate evaluation of multivariate candidate surrogates for binary clinical endpoints.  Our proposed method uses a semi-parametric estimated maximum likelihood fit for a Weibull time-to-event model.

@@

The dimension of the marginal-CEP versus candidate PS curve increases with the dimension of the candidate surrogate, making the marginal-CEP curve difficult to display and interpret for multivariate candidate PS. If instead we consider the clinical outcome within levels of the marginal-CEP, we can define useful curves that are two-dimensional regardless of the number of biomarkers included in the risk model. Our proposed curves include the marginal-CEP-based and time-dependent true positive fraction ($TPF(t|c)$) and false positive fraction ($FPF(t|c)$) as well as the receiver operating characteristic curve ($ROC^{t}(q)$) they define. We also consider the cumulative distribution function (CDF) of the marginal CEP function.

@@

\section{Notation and Criterion for Principal Surrogacy} \label{DEF}
Let $Z$ be the treatment indicator, 0 for control/non-active treatment and 1 for treatment. Let $W$ be a vector of baseline measurements taken prior to randomization. In the principal stratification framework of \citet{Frangakis02} we use potential outcomes, where all post-randomization measures are considered under assignment to either treatment arm for each individual. Let $T_i(z)$ be the potential time from randomization to clinical event for individual $i$ had s/he received treatment $z\in \{0,1\}$. Let $Y(z)$ be the indicator of $T(z)<C(z)$, where $C(z)$ is the potential censoring time and let $X(z)=min(T(z),C(z))$. Let there be $J$ candidate surrogates of interest which is measured prior to the clinical event but after randomization; with $S_{j}(z)$ being the $j$th candidate surrogate under treatment arm $z\in \{0,1\}$. Let $Y_{i}^{t}(z)$ be the indicator that $T_i(z)\leq t$. Let $D_i(t)=Y_{i}^{t}(0)-Y_{i}^{t}(1)$ be the individual treatment effect for subject $i$ on the clinical endpoint at or before time $t$. Let $\{S_{1,i}\ldots,S_{J,i}, T_i, C_i, Y_i, Y_{i}^{t} \}$ denote the observed values of the potential outcomes for subject $i$.

@@

\citet{Frangakis02} developed the concept of principal surrogacy and defined a PS by a criterion based on joint risk estimands, which conditions on the candidate surrogate under active treatment and under control, $risk_z(s_1,s_0)$. If $risk_1(s_1,s_0)=risk_0(s_1,s_0)$ for all $s_1 = s_0$, S is a principal surrogate; \citet{Gilbert08} called this average causal necessity (ACN). Subsequent work added a second criterion that a good PS should strongly modify the clinical treatment effects over the $(s_1, s_0)$ subgroups, the wide variability criterion. This can be demonstrated by investigating contrasts of $risk_1(s_1,s_0)$ and $risk_0(s_1,s_0)$, such as the ratio, over the levels of $(s_1, s_0)$. ACN is a sensible condition that can help eliminate candidate PS, but is not a sufficient condition, as was shown in \citet{Vanderweele11}. The combination of wide effect modification and ACN are often used for evaluation in the current literature. The wide effect modification criterion may also suggest a possible ranking of candidate surrogates based on the level of treatment efficacy modification over the subgroups defined by the surrogate.

In \citet{Gilbert08} and most of the recent literature, marginal risk estimands are used instead of joint risk estimands because they are easier to identify and estimate. marginal risk estimands condition on $S(1)$ and marginalize over the distribution of $S(0)$. The concept of wide variability is easily extended to marginal risk, such that contrasts of risk over the arms of the trial, such as our CEP of risk difference, vary widely in $s_1$. This can simply be viewed as a subgroup analysis, where subgroups defined by levels of $S(1)$, should strongly modify the clinical treatment effect. Other than the special case when the candidate surrogate is constant in the control arm, $S(0)=C$ called Constant Biomarker (CB), ACN cannot be verified based on the marginal risks alone \citep{Gilbert08, Gabriel13}. \citet{Gabriel13} outlines some PS candidate constructions that can make CB more likely to hold in a general clinical trial where biomarkers under control will tend not to be constant.

For a univariate candidate PS ACN can be displayed using the joint CEP, or marginal CEP under case CB.  Although the concept of ACN is easily extended to a multivariate PS, $risk_1(s_{1,1}, \dots, s_{1,J} ,s_{0,1}, \dots, s_{0,J})=risk_0(s_{1,1}, \dots, s_{1,J} ,s_{0,1} \dots s_{0,J})$ for all $s_{1,j} = s_{0,j}$ with $j \in \{1, \dots, J\}$, it is a more difficult concept to display graphically. As well, the definition does not seem as sensible in the multi-dimensional case. Clearly, we do not want a large amount of clinical treatment effect for a group of subjects with no treatment effect on any of the surrogates. However, ACN may not guarantee this is the case \citep{Vanderweele11}. We propose to extend the concept of ACN to investigate the ability of the model for risk difference conditional on a set of biomarkers to correctly categorize subjects by their treatment effect at time $t$, $D(t)$. For this purpose, we propose the use of a set of classification accuracy measures, outlined in Sections \ref{CAM} and \ref{CDF}. Like ACN, the classification accuracy measures allow for the quantification of the agreement between the treatment effect on the clinical outcome and the predictive model when there is no treatment effect on any of the biomarkers in the model.  Unlike ACN, the measures allow for quantification of this agreement for all other levels of the biomarkers as well.  These accuracy measures can also be displayed as curves versus the levels of the predictive model to allow for 2-dimensional  illustration of the predictive accuracy of the risk model. We suggest that good agreement between the predictive model containing the set of biomarkers and the clinical treatment efficacy is necessary but not sufficient for a set of biomarkers to have value as a PS. Much like ACN, it is possible to have good agreement between the model prediction and the treatment efficacy simply because there is no efficacy.

 For a multivariate PS the concept of wide variation in the CEP over subgroups defined by the set of biomarkers is clear. However, this is again difficult to display graphically due to the increased dimension of the surrogate. We propose the use of functions based on the CEP rather than the set of biomarker values for display of the wide variation criterion. We propose the CDF of the CEP function described in Section \ref{CDF} for graphical display of multivariate wide variation. We also suggest a summary statistic, the standardized total gain, that quantifies both our suggested accuracy concept and wide variation in a single measurement. Details for standardized total gain are given in Section \ref{STG}.  As well, if one assumes a parametric model for risk conditional on a set of biomarkers then a test for the minimum amount variation in the CEP over subgroups defined by the biomarkers can be based on the coefficients of the biomarkers in the model. The specifics of this test for our assumed model are given in Section \ref{CRE}.

@@

In order to link these risk estimands to the observable data we must make some assumptions; Assumptions A1-A4 reduce the number of missing potential outcomes and help identify the risk estimands.
\begin{itemize}
\item A1: Stable Unit Treatment Value Assumption (SUTVA) and Consistency 
\item A2: Ignorable Treatment Assignment
\item A3: Equal clinical risk up to time $\tau$: $T(1) \leq\tau$ if and only if $T(0) \leq\tau$
\item A4: Random censoring: $T(z) \perp C(z)$ for $z\in \{0,1\}$.
\end{itemize} 
Assumption A1 implies that subjects' potential outcomes are independent of other subjects' outcomes and treatment, SUTVA, and observation of the potential outcome does not change its value, consistency. For example, $(S_i|z_i=0) \equiv S_i(0)$ and this is independent of $(S_j|z_j=1) \equiv S_j(1)$, for $i \neq j$. SUTVA may not hold is some situations, such as cluster randomized vaccine trials where herd immunity may change the treatment effect on all outcomes in the cluster.  Assumption A2 implies that the distribution of the potential outcomes S(z) and Y(z) are independent of the observed treatment assignment. Assumption A2, is generally true in randomized, blinded trials.  Assumption A3 aides in the identification of subjects with $T < \tau$ to be excluded from the evaluation cohort without causing bias. Although, A3 is not fully testable, indications of A3 violation are significantly unbalanced event rates over the trial arms prior to $\tau$. Assumption A4 is standard in much of the survival analysis literature and allows for classic censoring models to be used. Assumptions A1-A4 imply that the conditional distribution of $T(z)$ given $\{S_{1}(1)=s_{1,1},\ldots,S_{J}(1)=s_{1,J}, T(1)\geq\tau, T(0)\geq\tau\}$, equals that for the observed outcome T given the observed treatment and potential candidate surrogate measures $\{Z=z, S_{1}(1)=s_{1,1},\ldots,S_{J}(1)=s_{1,J},T\geq\tau\}$ for $z\in \{0,1\}$.

@@

One can test the null hypothesis of no surrogate value or no variation in the marginal-CEP, $H0_1$, via the estimated model coefficients by a Wald test of the null hypothesis $\gamma_{0j}=\gamma_{1j}=0$ $\forall j \in \{1,\ldots J\}$. If any of the $S_{j}(1)$ associated terms significantly differ from zero then there is evidence of variation in the risk difference over the candidate PS.

@@

This likelihood could be solved by direct optimization if we assumed a parametric model for $S_j(1)|W$. However, as proposed in \citet{Pepe91}, consistent estimation of outcome model parameters can be achieved using an estimated distribution of $S_j(1)|W$; this is called estimated maximum likelihood. Following \citet{Huang11}, we assume the semi-parametric location-scale model of \citet{Heagerty99} for $F_{S_{j}(1)|W}$. Considering first a univariate candidate surrogate, the semi-parametric model is defined as $F_{S_{j}(1)|W} \sim F_j[\{s_{1,j} -\mu_j(w)\} / \sigma_j(w)]= F_j(\varsigma_j)$, where $F_j$ is the CDF of the univariate residuals $\varsigma_j$ from fitting the location and scale parameters $\mu_j(w)$ and $\sigma_j(w)$, respectively.  We define $\mu_j(w)$ and $\sigma_j(w)$ as parametric functions of the baseline variable(s) $W$, $\mu_j(w) = \gamma_j'w$ and $\log\{\sigma_j(w)\} = \eta_j'w$. We can estimate these parameters by solving the equations given in \citet{Heagerty99}:
\begin{eqnarray}
&&\sum_{k=1}^{n_V} {w_{k} (s_{(1,j,k)} - \gamma_j' w_{k} )\over \{\exp(\eta_j'w_k)\}^2} =0\\ \label{stuff}
&&\sum_{k=1}^{n_V} {w_{k}\{(s_{(1,j,k)} - \gamma_j' w_{k})^2 - \{\exp(\eta_j'w_k)\}^2\} \over \{\exp(\eta_j'w_k)\}^2}=0,\label{stuff2}
\end{eqnarray}
among those treated subjects with all $S_{(j,k)}(1)$ and $W$ measured. We denote the number of subjects in this validation group by $n_V$.

@@

In this case, sensitivity and specificity are based on unobservable data and cannot be estimated empirically. However, under assumptions A1-A5 plus the additional assumption of monotonicity of treatment effect, we can estimate them. Monotonicity of treatment effect is formally stated as
\begin{itemize}
\item A6: $T(z) \leq T(1-z)$ for one of the treatment arms $z$, $z\in \{0,1\}$.
\end{itemize}
Assumption A6 is needed to identify the classification accuracy measures. Assumption A6 is not fully testable, although it can be rejected. Given Assumptions A1-A5, there are clear testable implications of Assumption A6, such as $P(T\leq t|Z=1)\geq P(T\leq t|Z=0)$, and no switch in sign of the model-based estimate of risk difference over the range of the observed values of the $S_j(1)$s.

Assumption A6 may be more plausible in settings where the outcome is a specific illness or infection rather than death from any cause. Assuming A1-A6, $TPF(t|c)$ can be identified and estimated as we define it by:
\begin{eqnarray*}
TPF(t|c)&=& E\{\Delta(t|S_{1,1}, \dots, S_{1,J})\mathrm{I}_{[\Delta(t|S_{1,1}, \dots, S_{1,J})>c]}\} \over E\{\Delta(t|S_{1,1}, \dots, S_{1,J})\}.
\end{eqnarray*}
The derivation of this form is given in Appendix A of the supplementary materials; both the definition and derivation are adaptations of time-independent curves given in \citet{Huang12b}. The function $FPR(t|c)$ can be defined similarly based on the expectation of $1-\Delta(t|s_{1,1}, \dots, s_{1,J})$. One can estimate these accuracy measures using the plug-in estimators:
\begin{eqnarray*}
\widehat{TPF}(t|c)&=& {\int \hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})\mathrm{I}_{\{\hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})>c\}}d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J}) \over \int \hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J})} \hspace{2mm}\mbox{and},\\
\widehat{FPF}(t|c)&=& {\int 1-\hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})\mathrm{I}_{\{\hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})>c\}}d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J}) \over \int 1-\hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J})}.\\
\end{eqnarray*}
Here $\hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})$ is a maximum likelihood estimate under A1-A6 and using the estimated coefficients from the assumed model A5.

@@

When A6 is observably violated the suggested estimators $FPF(t|c)$ and $TPF(t|c)$ are biased. However, under minor violations of A6 that might not be observable in the sample, the estimators are still reasonable estimates of useful quantities. Greater discussion of this point is given in Appendix A of the supplementary materials. When there are obvious violations of A6, the estimation of $FPF(t|c)$ and $TPF(t|c)$ should not be pursued. When Assumptions A6 is clearly violated we suggest the use of the $\widetilde{STG}$ and $CDF^{t}_{\Delta}(c)$ for comparison and evaluation of candidate surrogates.

@@

\subsection{Integrated Standardized Total Gain} \label{ISTG}
As an alternative to estimating the $STG$ at a fixed time $t$, we can summarize the $STG(t)$ by averaging over the estimated distribution of event times. We define the integrated $STG$ ($ISTG$) as $\int STG(t) \, dF(t)$, where $F(t)$ is the marginal CDF of T. The estimated marginal distribution of event times $\hat{F}(T)$ can be calculated in a number of ways, by integrating over $Z$ and $S$ in the conditional CDF defined by the assumption A5 or Kaplan-Meier estimator. Then the plug-in estimator of the integrated $STG$ is the expectation with respect to that distribution:
\[
 \widetilde{STG} = \int \widehat{STG}(t) \, d \hat{F}(t). 
\]
The $\widetilde{STG}$ can be used to summarize the surrogate quality without specifying a fixed time point. By borrowing information across times, it provides a useful, single value estimate for describing and comparing candidate surrogates. The $\widetilde{STG}$ difference between the candidate PS in the same trial can be used in the same way as the difference in $STG(t)$ for inference on comparison.
